---
title: "Birding stats"
date: "2023-05-11"
draft: false
subtitle: ""
output: blogdown::html_page
excerpt: An attempt to transfer my childhood obsession with sport statistics into the one sport I engage in nowadays.
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

library(tidyverse)
library(here)
library(glue)
library(kableExtra)

i_am("content/kartrix/eBird/index.Rmd")

cur_path <- "content/kartrix/eBird/"


knitr::opts_chunk$set(echo = FALSE, include=FALSE, warnings = FALSE, messages = FALSE)

```

Birding to me is a lot of things; at times, I see it as a sport. Given my love for numbers and my obsession with cricket stats as a kid (more on this coming soon!), I wanted to try and implement a similar sort of analysis and visualisation of sport statistics.

**Disclaimer:** All this is based off my data uploaded to eBird. That covers most, but not all, of the birding I have done.

# My career in numbers {#numbers}

```{r}

rawdata <- read.csv(here(glue("{cur_path}MyEBirdData.csv"))) %>% 
  # removing unnecessary columns and renaming necessary ones
  dplyr::select(-c(Scientific.Name, Area.Covered..ha., ML.Catalog.Numbers)) %>% 
  # for uniformity with regular EBD
  magrittr::set_colnames(c("SAMPLING.EVENT.IDENTIFIER", "COMMON.NAME", "TAXONOMIC.ORDER",
                           "OBSERVATION.COUNT", "STATE.CODE", "COUNTY", 
                           "LOCALITY.ID", "LOCALITY", "LATITUDE", "LONGITUDE", 
                           "OBSERVATION.DATE", "TIME.OBSERVATION.STARTED", "PROTOCOL.TYPE",
                           "DURATION.MINUTES", "ALL.OBS.REPORTED", "EFFORT.DISTANCE.KM",
                           "NUMBER.OBSERVERS", "BREEDING.CODE", 
                           "SPECIES.COMMENTS", "TRIP.COMMENTS"))

# remove spuhs and slashes
nonspec <- rawdata %>% 
  filter(str_ends(COMMON.NAME, "sp.") |
           str_detect(COMMON.NAME, "/")) %>% 
  # ignoring slashes in true species
  filter(!(COMMON.NAME %in% c("Long-tailed Shrike (tricolor/longicaudatus)",
                              "Western Yellow Wagtail (flava/beema)"))) %>% 
  distinct(COMMON.NAME)

rawdata <- rawdata %>% anti_join(nonspec, by = "COMMON.NAME")

tot_lists <- n_distinct(rawdata$SAMPLING.EVENT.IDENTIFIER)

# 50s, 100s, average score
stats0 <- rawdata %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  dplyr::summarise(NO.SP = n_distinct(COMMON.NAME)) %>% 
  mutate(AVERAGE.SCORE = mean(NO.SP) %>% floor()) %>% 
  filter(NO.SP >= 50) %>% 
  mutate(`50s` = n_distinct(SAMPLING.EVENT.IDENTIFIER)) %>% 
  filter(NO.SP >= 100) %>% 
  dplyr::summarise(`50s` = min(`50s`),
                   `100s` = n_distinct(SAMPLING.EVENT.IDENTIFIER),
                   `Average score` = min(AVERAGE.SCORE)) %>% 
  pivot_longer(everything(), names_to = "STATISTIC", values_to = "VALUE") %>% 
  mutate(UNIT = c("-", "-", "spp."),
         CHECKLIST = "-",
         VALUE = glue("{VALUE}"))

# all-time high score
high_score <- rawdata %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER, DURATION.MINUTES, STATE.CODE) %>% 
  dplyr::summarise(NO.SP = n_distinct(COMMON.NAME)) %>% 
  ungroup() %>% 
  filter(NO.SP == max(NO.SP)) %>% 
  transmute(STATISTIC = "Highest score",
            VALUE = glue("{NO.SP} ({DURATION.MINUTES}) in {STATE.CODE}"),
            UNIT = "spp. (min)",
            CHECKLIST = glue("https://ebird.org/checklist/{SAMPLING.EVENT.IDENTIFIER}"))

# high score in 1 hour
high_score_hour <- rawdata %>% 
  filter(DURATION.MINUTES <= 60) %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER, DURATION.MINUTES, STATE.CODE) %>% 
  dplyr::summarise(NO.SP = n_distinct(COMMON.NAME)) %>% 
  ungroup() %>% 
  filter(NO.SP == max(NO.SP)) %>% 
  transmute(STATISTIC = "Highest score within 1 hour",
            VALUE = glue("{NO.SP} ({DURATION.MINUTES}) in {STATE.CODE}"),
            UNIT = "spp. (min)",
            CHECKLIST = glue("https://ebird.org/checklist/{SAMPLING.EVENT.IDENTIFIER}"))

# high and low speeds
stats_speed <- rawdata %>% 
  filter(ALL.OBS.REPORTED == 1,
         # removing lists without duration info
         !is.na(DURATION.MINUTES)) %>% 
  group_by(SAMPLING.EVENT.IDENTIFIER) %>% 
  dplyr::summarise(NO.SP = n_distinct(COMMON.NAME),
                   DURATION.MINUTES = min(DURATION.MINUTES)) %>% 
  mutate(SPEED = round((NO.SP/DURATION.MINUTES), 3), # species per hour
         HIGH.SPEED = max(SPEED),
         LOW.SPEED = min(SPEED)) %>% 
  filter(SPEED == HIGH.SPEED | SPEED == LOW.SPEED) %>% 
  # column saying high or low
  mutate(STATISTIC = if_else(SPEED == HIGH.SPEED, 
                             "Highest speed", 
                             "Lowest speed")) %>% 
  left_join(rawdata, by = c("SAMPLING.EVENT.IDENTIFIER", "DURATION.MINUTES")) %>% 
  distinct(SPEED, STATISTIC, STATE.CODE, SAMPLING.EVENT.IDENTIFIER) %>% 
  arrange(desc(SPEED)) %>% 
  group_by(SPEED, STATISTIC) %>% 
  reframe(CHECKLIST = glue("https://ebird.org/checklist/{SAMPLING.EVENT.IDENTIFIER}"),
          VALUE = glue("{SPEED} ({SPEED*60}) in {STATE.CODE}")) %>% 
  group_by(STATISTIC) %>% 
  dplyr::summarise(VALUE = str_flatten_comma(VALUE),
                   UNIT = "spp./min (spp./h)",
                   CHECKLIST = str_flatten_comma(CHECKLIST))

# combining all
stats_all <- stats0 %>% 
  bind_rows(high_score, high_score_hour, stats_speed)
  
# # to decide how wide cell should be
# wrap_width_value <- str_split(stats_all$VALUE, ", ") %>% 
#   map_dbl(~ str_length(.) %>% max()) %>% 
#   max()
# wrap_width_checklist <- str_split(stats_all$CHECKLIST, ", ") %>% 
#   map_dbl(~ str_length(.) %>% max()) %>% 
#   max()

```

```{r, include=TRUE}

table_caption <- glue("My birding stats, as analysed from {tot_lists} checklists. This only includes non&ndash;zero-species checklists, and this directly impacts metrics such as lowest speed.")

# tables of stats
stats_all %>% 
  mutate(across(everything(), ~ as.character(.))) %>% 
  knitr::kable(caption = table_caption, booktabs = TRUE) %>%
  # first column should be bold
  column_spec(1, bold = TRUE, width_min = "9em") %>%
  # adjusting cell widths
  column_spec(2, width_min = "12em") %>%
  column_spec(4, width_min = "19em")

```


# My career mapped {#map}

Spatial visualisation of my eBirding in the style of an eBird profile, but as a bigger sandbox... **Coming soon!**
